---
title: "Comparative inference"
---

## General utils
```{r}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


```{r warning=FALSE}
library(compositions)
library(DepthProc)

library(ggtern)
library(RColorBrewer)
library(lattice)
library(wesanderson)
```


Loading the data
```{r}
set.seed(050700)

root = "../.."
source(paste(root,"include/utils.R",sep = "/"))

library(readr)
clean_data <- read_csv(paste(root, "data/clean_merged_data.csv", sep="/"))
```
### Inference functions
```{r}
# set of functions to perform non-parametric tests
source("inference_module.R")

B <- 1e4
alpha <- 0.05
```

Here we want to inspect if there are differences between the distribution of some variables in "republican" and "democratic" counties
```{r}
gop_counties <- which(clean_data$gop_pct20 > clean_data$dem_pct20)
dem_counties <- which(clean_data$gop_pct20 < clean_data$dem_pct20)
```

We're gonna consider the following test statistics
```{r}
mult_mean_diff <- function(df1, df2){
  m1 <- apply(df1,2, mean)
  m2 <- apply(df2,2, mean)
  
  cov1 <- stats::cov(df1)
  cov2 <- stats::cov(df2)
  
  n1 <- nrow(df1)
  n2 <- nrow(df2)
  
  return(as.numeric(t(m1-m2) %*% solve(1/n1*cov1 + 1/n2*cov1) %*% (m1-m2)))
}

mult_median_diff <- function(df1, df2){
  library(DepthProc)
  
  m1 <- DepthProc::depthMedian(df1)
  m2 <- DepthProc::depthMedian(df2)
  
  return(max(abs(m1 - m2)))
}
```


## Ethnical composition
Mapping the composition to an euclidean space
```{r}
race.composit <- acomp(clean_data, 
                   parts = c("white_pct", "black_pct", "hispanic_pct"))

mapped_race.composit <- ilr(race.composit)
```

### DD-plot
```{r}
ddPlot(x = as.data.frame(scale(mapped_race.composit[dem_counties,])), 
       y = as.data.frame(scale(mapped_race.composit[gop_counties,])),
       depth_params = list(method='Tukey'))
```


### Bootstrap estimation
Estimation
```{r}
# permutational parameters
alpha <- 0.05
k <- 2

B <- 1e4
seed <- 050700

n_cores <- 8
cl <- makeCluster(n_cores)

# computation of the medians by county majority
dem_race.med.estimation <- boot.estimation(mapped_race.composit[dem_counties,],
                                              DepthProc::depthMedian, 
                                              alpha/k, 
                                              B, seed, cl)
dem_race.median <- ilrInv(dem_race.med.estimation$obs.T_stat)
names(dem_race.median) <- names(race.composit)

gop_race.med.estimation <- boot.estimation(mapped_race.composit[gop_counties,],
                                              DepthProc::depthMedian, 
                                              alpha/k, 
                                              B, seed, cl)
gop_race.median <- ilrInv(gop_race.med.estimation$obs.T_stat)
names(gop_race.median) <- names(race.composit)
```

Plotting
```{r}
# little utility function to plot results from the bootstrap estimation
estimation_report <- function(dem_median_estimation, gop_median_estimation, 
                              mapped.composit,
                              groups_names){
  dem_median <- ilrInv(dem_median_estimation$obs.T_stat)
  names(dem_median) <- groups_names
  
  gop_median <- ilrInv(gop_median_estimation$obs.T_stat)
  names(gop_median) <- groups_names
  
  #barplot comparing the medians
  barplot(cbind(dem_median, gop_median), legend.text = T, 
          col = wes_palette(n = length(groups_names), name = "GrandBudapest2"))
  
  #pair plot in the space of the compositions
  splom(as.matrix(mapped.composit), 
      groups=ifelse(clean_data$dem_pct20 > clean_data$gop_pct20, 1, 0), 
      pch=21,
      panel=function(x, y, i, j, groups, ...) {
                panel.points(
                  x, y, pch=20, cex = 0.2,
                  col=ifelse(clean_data$dem_pct20 > clean_data$gop_pct20,
                             "cornflowerblue", "coral"))
                panel.points(dem_median_estimation$obs.T_stat[j],
                             dem_median_estimation$obs.T_stat[i],
                             pch=".",col='darkblue',cex=1)
                panel.points(gop_median_estimation$obs.T_stat[j], 
                             gop_median_estimation$obs.T_stat[i],
                             pch=".",col='firebrick',cex=1)
                panel.rect(xleft = dem_median_estimation$CI.RP[1,j],
                           xright = dem_median_estimation$CI.RP[2,j],
                           ybottom = dem_median_estimation$CI.RP[1,i],
                           ytop = dem_median_estimation$CI.RP[2,i],
                           col = "darkblue")
                panel.rect(xleft = gop_median_estimation$CI.RP[1,j],
                           xright = gop_median_estimation$CI.RP[2,j],
                           ybottom = gop_median_estimation$CI.RP[1,i],
                           ytop = gop_median_estimation$CI.RP[2,i],
                           col = "firebrick")
            })
}
```

```{r}
estimation_report(dem_race.med.estimation, gop_race.med.estimation,
                  mapped_race.composit,
                  names(race.composit))
```

### t-test
Permutational test
```{r}
perm_test_race <- perm_t.test(
  mapped_race.composit[dem_counties,], mapped_race.composit[gop_counties,],
  mult_median_diff,
  B, seed, cl
)

plot.perm_results(perm_test_race)
```

## Age composition
```{r}
age.composit <- acomp(clean_data, 
                  parts = c("age29andunder_pct", "age30to64", "age65andolder_pct")
                  )
mapped_age.composit <- ilr(age.composit)
```

### DD-plot
```{r}
ddPlot(x = as.data.frame(scale(mapped_age.composit[dem_counties,])), 
       y = as.data.frame(scale(mapped_age.composit[gop_counties,])),
       depth_params = list(method='Tukey'))
```


### Bootstrap estimation
```{r}
# permutational parameters
alpha <- 0.01
k <- 2

B <- 1e4
seed <- 050700

n_cores <- 8
cl <- makeCluster(n_cores)

# computation of the medians by county majority
dem_age.med.estimation <- boot.estimation(mapped_age.composit[dem_counties,],
                                              DepthProc::depthMedian, 
                                              alpha/k, 
                                              B, seed, cl)
dem_age.median <- ilrInv(dem_age.med.estimation$obs.T_stat)
names(dem_age.median) <- names(age.composit)

gop_age.med.estimation <- boot.estimation(mapped_age.composit[gop_counties,],
                                              DepthProc::depthMedian, 
                                              alpha/k, 
                                              B, seed, cl)
gop_age.median <- ilrInv(gop_age.med.estimation$obs.T_stat)
names(gop_age.median) <- names(age.composit)
```

Plotting
```{r}
estimation_report(dem_age.med.estimation, gop_age.med.estimation,
                  mapped_age.composit,
                  names(age.composit))
```

Tern plots
```{r}
library(ggtern)

# rectangle confidence intervals in the mapped space of compositions
dem.rect_vert <- with(dem_age.med.estimation,
                  data.frame(A = c(CI.RP[1,1],CI.RP[1,2]),
                             B = c(CI.RP[1,1],CI.RP[2,2]),
                             C = c(CI.RP[2,1],CI.RP[2,2]),
                             D = c(CI.RP[2,1],CI.RP[1,2])
                  )
)

gop.rect_vert <- with(gop_age.med.estimation,
                  data.frame(A = c(CI.RP[1,1],CI.RP[1,2]),
                             B = c(CI.RP[1,1],CI.RP[2,2]),
                             C = c(CI.RP[2,1],CI.RP[2,2]),
                             D = c(CI.RP[2,1],CI.RP[1,2])
                  )
)

# visualization
ggtern() +
  geom_point(aes(x=age29andunder_pct, y=age30to64, z=age65andolder_pct),
             data = age.composit,
             color = ifelse(clean_data$dem_pct20 > clean_data$gop_pct20,
                             "cornflowerblue", "coral")) +
  geom_polygon(aes(x=age29andunder_pct, y=age30to64, z=age65andolder_pct),
               data = ilrInv(t(dem.rect_vert), orig = age.composit),
               color = 'darkblue') +
  geom_polygon(aes(x=age29andunder_pct, y=age30to64, z=age65andolder_pct),
               data = ilrInv(t(gop.rect_vert), orig = age.composit),
               color = 'darkred') +
  theme_bw()
```
```{r}
plot(age.composit, 
     col = ifelse(clean_data$dem_pct20 > clean_data$gop_pct20,
                             "cornflowerblue", "coral"),
     pch=20, cex = 0.1,
     axes=T, labels= c("under30","30to64","over64"))

with(gop.rect_vert,
     list(
       segments.acomp(ilrInv(A),ilrInv(B), col="darkred"),
       segments.acomp(ilrInv(B),ilrInv(C), col="darkred"),
       segments.acomp(ilrInv(C),ilrInv(D), col="darkred"),
       segments.acomp(ilrInv(D),ilrInv(A), col="darkred")
     )
)

with(dem.rect_vert,
     list(
       segments.acomp(ilrInv(A),ilrInv(B), col="darkblue"),
       segments.acomp(ilrInv(B),ilrInv(C), col="darkblue"),
       segments.acomp(ilrInv(C),ilrInv(D), col="darkblue"),
       segments.acomp(ilrInv(D),ilrInv(A), col="darkblue")
     )
)

plot(ilrInv(gop_age.med.estimation$obs.T_stat), col = "red", add=T, pch=8, cex = 0.5)
plot(ilrInv(dem_age.med.estimation$obs.T_stat), col = "blue", add=T, pch=8, cex = 0.5)

```


### t-test
Permutational test
```{r}
perm_test_age <- perm_t.test(
  mapped_age.composit[dem_counties,], mapped_age.composit[gop_counties,],
  mult_median_diff,
  B, seed, cl
)

plot.perm_results(perm_test_age)
```

## Gender composition
```{r}
gender.composit <- acomp(clean_data, 
                     parts = c("male_pct", "female_pct"))
mapped_gender.composit <- ilr(gender.composit)
```

### QQ-plot
We're in the univariate case, thus just a qqplot is needed
```{r}
qqplot(x = scale(mapped_gender.composit[dem_counties,1]), 
       y = scale(mapped_gender.composit[gop_counties,1]),
       xlim = c(-10,10), ylim = c(-10,10)
       )
abline(a = 0, b = 1, col = "red")
```

### Bootstrap estimation
```{r}
#bootstrap parameters
alpha <- 0.05
k <- 1

B <- 1e4
seed <- 050700

n_cores <- 8
cl <- makeCluster(n_cores)

med.t_stat <- function(x){
  median(as.numeric(x))
}

# computation of the medians by county majority
dem_gender.med.estimation <- boot.uni.estimation(mapped_gender.composit[dem_counties,],
                                              med.t_stat, 
                                              alpha/k, 
                                              B, seed, cl)
dem_gender.median <- ilrInv(dem_gender.med.estimation$obs.T_stat)
names(dem_gender.median) <- names(gender.composit)

gop_gender.med.estimation <- boot.uni.estimation(mapped_gender.composit[gop_counties,],
                                              med.t_stat, 
                                              alpha/k, 
                                              B, seed, cl)
gop_gender.median <- ilrInv(gop_gender.med.estimation$obs.T_stat)
names(gop_gender.median) <- names(gender.composit)
```

Plotting
```{r}
#barplot comparing the medians
barplot(cbind(dem_gender.median, gop_gender.median), legend.text = T, 
          col = wes_palette(n = length(names(gender.composit)), name = "GrandBudapest2"))

dem_gender.med.estimation$CI.RP
gop_gender.med.estimation$CI.RP
```
### t-test
Permutational test
```{r}
perm_test_gender <- perm_t.uni.test(
  mapped_gender.composit[dem_counties,1], mapped_gender.composit[gop_counties,1],
  function(data1,data2){abs(median(data1)-median(data2))},
  B, seed, cl)

plot.perm_results(perm_test_gender)
```

## Educational composition
Mapping the composition to an euclidean space
```{r}
edu.composit <- acomp(clean_data, 
                   parts = c("lesshs_pct", "onlyhs_pct", "morecollege_pct"))
mapped_edu.composit <- ilr(edu.composit)
```

### DD-plot
```{r}
ddPlot(x = as.data.frame(scale(mapped_edu.composit[dem_counties,])), 
       y = as.data.frame(scale(mapped_edu.composit[gop_counties,])),
       depth_params = list(method='Tukey'))
```

### Bootstrap estimation
```{r}
# permutational parameters
alpha <- 0.05
k <- 3

B <- 1e4
seed <- 050700

n_cores <- 8
cl <- makeCluster(n_cores)

# computation of the medians by county majority
dem_edu.med.estimation <- boot.estimation(mapped_edu.composit[dem_counties,],
                                              DepthProc::depthMedian, 
                                              alpha/k, 
                                              B, seed, cl)
dem_edu.median <- ilrInv(dem_edu.med.estimation$obs.T_stat)
names(dem_edu.median) <- names(edu.composit)

gop_edu.med.estimation <- boot.estimation(mapped_edu.composit[gop_counties,],
                                              DepthProc::depthMedian, 
                                              alpha/k, 
                                              B, seed, cl)
gop_edu.median <- ilrInv(gop_edu.med.estimation$obs.T_stat)
names(gop_edu.median) <- names(edu.composit)
```

Plotting
```{r}
estimation_report(dem_edu.med.estimation, gop_edu.med.estimation,
                  mapped_edu.composit, 
                  names(edu.composit))
```

### t-tests
Permutational test
```{r}
perm_test_edu <- perm_t.test(
  mapped_edu.composit[dem_counties,], mapped_edu.composit[gop_counties,],
  mult_median_diff,
  B, seed, cl
)

plot.perm_results(perm_test_edu)
```

## Whites edu composition
Mapping the composition to an euclidean space
```{r}
comp_eduW <- acomp(clean_data, 
                   parts = c("lesshs_pct", "onlyhs_pct", "morecollege_pct"))
mapped_eduW_composit <- ilr(comp_edu)
```

### Bootstrap estimation
```{r}
alpha <- 0.05
k <- 2 #correction

depth_median <- function(x){
  return(DepthProc::depthMedian(x))
}

# computation of the medians by county majority
dem_eduW_median_estimation <- boot_estimation(mapped_eduW_composit[dem_counties,],
                                         depth_median, 
                                         alpha/k, B)
dem_eduW_median <- ilrInv(dem_eduW_median_estimation$estimated_t)
names(dem_eduW_median) <- names(comp_eduW)

gop_eduW_median_estimation <- boot_estimation(mapped_eduW_composit[gop_counties,],
                                         depth_median, 
                                         alpha/k, B)
gop_eduW_median <- ilrInv(gop_eduW_median_estimation$estimated_t)
names(gop_eduW_median) <- names(comp_eduW)
```

Plotting
```{r}
# comparison of the medians
barplot(cbind(dem_eduW_median, gop_eduW_median), legend.text = T, col = c("red", "blue", "green"))
```

```{r}
library(lattice)

splom(as.matrix(mapped_eduW_composit), 
      groups=ifelse(clean_data$gop_pct20 < clean_data$dem_pct20, 1, 0), pch=21,
      panel=function(x, y, i, j, groups, ...) {
                panel.points(x, y, pch=20, cex = 0.2,
                             col=ifelse(clean_data$gop_pct20 < 
                                          clean_data$dem_pct20, "cornflowerblue", "coral"))
                panel.points(dem_eduW_median_estimation$estimated_t[j],
                             dem_eduW_median_estimation$estimated_t[i],
                             pch=10,col='darkblue')
                panel.points(gop_edu_median_estimation$estimated_t[j], 
                             gop_edu_median_estimation$estimated_t[i],
                             pch=10,col='firebrick')
                panel.rect(xleft = dem_edu_median_estimation$CI.RP[1,j],
                           xright = dem_edu_median_estimation$CI.RP[2,j],
                           ybottom = dem_edu_median_estimation$CI.RP[1,i],
                           ytop = dem_edu_median_estimation$CI.RP[2,i],
                           col = "darkblue")
                panel.rect(xleft = gop_edu_median_estimation$CI.RP[1,j],
                           xright = gop_edu_median_estimation$CI.RP[2,j],
                           ybottom = gop_edu_median_estimation$CI.RP[1,i],
                           ytop = gop_edu_median_estimation$CI.RP[2,i],
                           col = "firebrick")
            })
```

### Tests
Permutational test
```{r}
perm_test_edu <- perm_T_test(
  as.data.frame(mapped_edu_composit), 
  gop_counties, dem_counties, 
  mult_median_diff,
  B)
```

```{r}
# PLOTS
#-> histograms of the permutational distribution
hist(perm_test_edu$perm_distr_T, 
     xlim=range(c(perm_test_edu$perm_distr_T, perm_test_edu$T0)))
abline(v=perm_test_edu$T0, col=3,lwd=4)
#-> distribution function of the permutational distribution
plot(ecdf(perm_test_edu$perm_distr_T), 
     xlim=range(c(perm_test_edu$perm_distr_T, perm_test_edu$T0)))
abline(v=perm_test_edu$T0, col=3,lwd=4)
  
# P-VALUE
p_val <- sum(perm_test_edu$perm_distr_T >= perm_test_edu$T0)/B
cat("p-value", p_val, "\n")
```

Bootstrapping test
```{r}
boot_test_edu <- boot_T_test(
  as.data.frame(mapped_edu_composit), 
  gop_counties, dem_counties, 
  mult_mean_diff,
  alpha, 10)
```


```{r}
# PLOTS
#-> histograms of the bootstrap distribution
hist(boot_test_edu$boot_distr_T, 
     xlim=range(c(boot_test_edu$boot_distr_T, boot_test_edu$T0)))
abline(v=boot_test_edu$T0, col=3,lwd=4)

#-> distribution function of the bootstrap distribution
plot(ecdf(boot_test_edu$boot_distr_T), 
     main='empirical bootstrap distr. of the t-test statistic')
abline(v=boot_test_edu$T0, lty=2)
abline(v = boot_test_edu$CI.RP, col='red')
abline(v = boot_test_edu$CI.P, col="blue")
```

## Wealth
### Bootstrap estimation
```{r}
alpha <- 0.05
k <- 2 #correction

depth_median <- function(x){
  return(DepthProc::depthMedian(x))
}

# computation of the medians by county majority
dem_income_median_estimation <- uni_boot_estimation(clean_data$median_hh_inc[dem_counties],
                                         median, 
                                         alpha/k, B)
gop_eduW_median_estimation <- uni_boot_estimation(clean_data$median_hh_inc[gop_counties],
                                         median, 
                                         alpha/k, B)
```


## For the report
- Ethnical composition
```{r}
race.composit <- acomp(clean_data, 
                   parts = c("white_pct", "black_pct", "hispanic_pct"))

mapped_race.composit <- ilr(race.composit)

race.dem_median <- ilrInv(dem_race.med.estimation$obs.T_stat, orig = race.composit)
race.gop_median <- ilrInv(gop_race.med.estimation$obs.T_stat, orig = race.composit)

## DD plots
png(filename = "plots/ethn_ddplot.png",
     width = 800, height = 800, units = "px")

ddPlot(x = as.data.frame(scale(mapped_race.composit[dem_counties,])), 
       y = as.data.frame(scale(mapped_race.composit[gop_counties,])),
       depth_params = list(method='Tukey'))

dev.off()

## Depth medians
png(filename = "plots/ethn_medians.png",
     width = 800, height = 800, units = "px")

barplot(cbind(race.dem_median, race.gop_median), 
          col = wes_palette(n = dim(race.composit)[2], name = "GrandBudapest2"),
        legend.text = c("white", "black", "hispanic"))

dev.off()


## Bootstrap confidence intervals
dem.rect_vert <- with(dem_race.med.estimation,
                  data.frame(A = c(CI.RP[1,1],CI.RP[1,2]),
                             B = c(CI.RP[1,1],CI.RP[2,2]),
                             C = c(CI.RP[2,1],CI.RP[2,2]),
                             D = c(CI.RP[2,1],CI.RP[1,2])
                  )
)

gop.rect_vert <- with(gop_race.med.estimation,
                  data.frame(A = c(CI.RP[1,1],CI.RP[1,2]),
                             B = c(CI.RP[1,1],CI.RP[2,2]),
                             C = c(CI.RP[2,1],CI.RP[2,2]),
                             D = c(CI.RP[2,1],CI.RP[1,2])
                  )
)

png(filename = "plots/race_boot_estimation.png",
     width = 800, height = 800, units = "px")

plot(race.composit, 
     col = ifelse(clean_data$dem_pct20 > clean_data$gop_pct20,
                             "cornflowerblue", "coral"),
     pch=20, cex = 0.1,
     plotMissings=F,
     axes=T, labels= c("white","black","hispanic"))

with(gop.rect_vert,
     list(
       segments.acomp(ilrInv(A),ilrInv(B), col="darkred"),
       segments.acomp(ilrInv(B),ilrInv(C), col="darkred"),
       segments.acomp(ilrInv(C),ilrInv(D), col="darkred"),
       segments.acomp(ilrInv(D),ilrInv(A), col="darkred")
     )
)

with(dem.rect_vert,
     list(
       segments.acomp(ilrInv(A),ilrInv(B), col="darkblue"),
       segments.acomp(ilrInv(B),ilrInv(C), col="darkblue"),
       segments.acomp(ilrInv(C),ilrInv(D), col="darkblue"),
       segments.acomp(ilrInv(D),ilrInv(A), col="darkblue")
     )
)

plot(ilrInv(gop_race.med.estimation$obs.T_stat), col = "red", add=T, pch=8, cex = 0.5)
plot(ilrInv(dem_race.med.estimation$obs.T_stat), col = "blue", add=T, pch=8, cex = 0.5)

dev.off()
```

```{r}
plot(age.composit, type="n")
points(race.composit, col="blue")   
```


- Educational composition
```{r}
edu.composit <- acomp(clean_data, 
                   parts = c("lesshs_pct", "onlyhs_pct", "morecollege_pct"))
mapped_edu.composit <- ilr(edu.composit)

edu.dem_median <- ilrInv(dem_edu.med.estimation$obs.T_stat, orig = edu.composit)
edu.gop_median <- ilrInv(gop_edu.med.estimation$obs.T_stat, orig = edu.composit)

## DD-plots
png(filename = "plots/edu_ddplot.png",
     width = 800, height = 800, units = "px")

ddPlot(x = as.data.frame(scale(mapped_edu.composit[dem_counties,])), 
       y = as.data.frame(scale(mapped_edu.composit[gop_counties,])),
       depth_params = list(method='Tukey'))

dev.off()

## Depth medians
png(filename = "plots/edu_medians.png",
     width = 800, height = 800, units = "px")

barplot(cbind(edu.dem_median, edu.gop_median), 
        col = wes_palette(n = dim(edu.composit)[2], name = "GrandBudapest2"),
        legend.text = c("less_hs","only_hs","more_hs"))

dev.off()

## Bootstrap confidence intervals 
dem.rect_vert <- with(dem_edu.med.estimation,
                  data.frame(A = c(CI.RP[1,1],CI.RP[1,2]),
                             B = c(CI.RP[1,1],CI.RP[2,2]),
                             C = c(CI.RP[2,1],CI.RP[2,2]),
                             D = c(CI.RP[2,1],CI.RP[1,2])
                  )
)

gop.rect_vert <- with(gop_edu.med.estimation,
                  data.frame(A = c(CI.RP[1,1],CI.RP[1,2]),
                             B = c(CI.RP[1,1],CI.RP[2,2]),
                             C = c(CI.RP[2,1],CI.RP[2,2]),
                             D = c(CI.RP[2,1],CI.RP[1,2])
                  )
)

png(filename = "plots/edu_boot_estimation.png",
     width = 800, height = 800, units = "px")

plot(edu.composit, 
     col = ifelse(clean_data$dem_pct20 > clean_data$gop_pct20,
                             "cornflowerblue", "coral"),
     pch=20, cex = 0.1,
     axes=T, labels= c("less_hs","only_hs","more_hs"))

with(gop.rect_vert,
     list(
       segments.acomp(ilrInv(A),ilrInv(B), col="darkred"),
       segments.acomp(ilrInv(B),ilrInv(C), col="darkred"),
       segments.acomp(ilrInv(C),ilrInv(D), col="darkred"),
       segments.acomp(ilrInv(D),ilrInv(A), col="darkred")
     )
)

with(dem.rect_vert,
     list(
       segments.acomp(ilrInv(A),ilrInv(B), col="darkblue"),
       segments.acomp(ilrInv(B),ilrInv(C), col="darkblue"),
       segments.acomp(ilrInv(C),ilrInv(D), col="darkblue"),
       segments.acomp(ilrInv(D),ilrInv(A), col="darkblue")
     )
)

plot(ilrInv(gop_edu.med.estimation$obs.T_stat), col = "red", add=T, pch=8, cex = 0.5)
plot(ilrInv(dem_edu.med.estimation$obs.T_stat), col = "blue", add=T, pch=8, cex = 0.5)

dev.off()
```

- Age compositions
```{r}
age.composit <- acomp(clean_data, 
                  parts = c("age29andunder_pct", "age30to64", "age65andolder_pct")
                  )
mapped_age.composit <- ilr(age.composit)

age.dem_median <- ilrInv(dem_age.med.estimation$obs.T_stat, orig = age.composit)
age.gop_median <- ilrInv(gop_age.med.estimation$obs.T_stat, orig = age.composit)

## DD-plots
png(filename = "plots/age_ddplot.png",
     width = 800, height = 800, units = "px")

ddPlot(x = as.data.frame(scale(mapped_age.composit[dem_counties,])), 
       y = as.data.frame(scale(mapped_age.composit[gop_counties,])),
       depth_params = list(method='Tukey'))

dev.off()

## Depth medians
png(filename = "plots/age_medians.png",
     width = 800, height = 800, units = "px")

barplot(cbind(age.dem_median, age.gop_median), 
          col = wes_palette(n = dim(age.composit)[2], name = "GrandBudapest2"),
        legend.text = c("under30", "30to64", "over64"))

dev.off()

## Bootstrap confidence intervals
dem.rect_vert <- with(dem_age.med.estimation,
                  data.frame(A = c(CI.RP[1,1],CI.RP[1,2]),
                             B = c(CI.RP[1,1],CI.RP[2,2]),
                             C = c(CI.RP[2,1],CI.RP[2,2]),
                             D = c(CI.RP[2,1],CI.RP[1,2])
                  )
)

gop.rect_vert <- with(gop_age.med.estimation,
                  data.frame(A = c(CI.RP[1,1],CI.RP[1,2]),
                             B = c(CI.RP[1,1],CI.RP[2,2]),
                             C = c(CI.RP[2,1],CI.RP[2,2]),
                             D = c(CI.RP[2,1],CI.RP[1,2])
                  )
)

png(filename = "plots/age_boot_estimation.png",
     width = 800, height = 800, units = "px")

plot(age.composit, 
     col = ifelse(clean_data$dem_pct20 > clean_data$gop_pct20,
                             "cornflowerblue", "coral"),
     pch=20, cex = 0.1,
     axes=T, labels= c("under30","30to64","over64"))

with(gop.rect_vert,
     list(
       segments.acomp(ilrInv(A),ilrInv(B), col="darkred"),
       segments.acomp(ilrInv(B),ilrInv(C), col="darkred"),
       segments.acomp(ilrInv(C),ilrInv(D), col="darkred"),
       segments.acomp(ilrInv(D),ilrInv(A), col="darkred")
     )
)

with(dem.rect_vert,
     list(
       segments.acomp(ilrInv(A),ilrInv(B), col="darkblue"),
       segments.acomp(ilrInv(B),ilrInv(C), col="darkblue"),
       segments.acomp(ilrInv(C),ilrInv(D), col="darkblue"),
       segments.acomp(ilrInv(D),ilrInv(A), col="darkblue")
     )
)

plot(ilrInv(gop_age.med.estimation$obs.T_stat), col = "red", add=T, pch=8, cex = 0.5)
plot(ilrInv(dem_age.med.estimation$obs.T_stat), col = "blue", add=T, pch=8, cex = 0.5)

dev.off()
```

